---
title: "Getting Started with nycOpenData"
output: rmarkdown::html_vignette
author: "Christian Martinez"
vignette: >
  %\VignetteIndexEntry{Getting Started with nycOpenData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(nycOpenData)
library(ggplot2)
library(dplyr)
```

## Introduction

Welcome to the `nycOpenData` package, a R package dedicated to helping R users connect to the [NYC Open Data Portal](https://opendata.cityofnewyork.us/)!

The `nycOpenData` package provides a streamlined interface for accessing New York City's vast open data resources. It connects directly to the NYC Open Data Portal, helping users bridge the gap between raw city APIs and tidy data analysis. It does this in two ways:

### The `nyc_pull_dataset()` function

The primary way to pull data in this package is the `nyc_pull_dataset()` function, which works in tandem with `nyc_list_datasets()`. Most workflows do not require users to manually construct API requests or manage authentication details.

The first step would be to call the `nyc_list_datasets()` to see what datasets are in the list and available to use in the `nyc_pull_dataset()` function. This provides information about datasets available through the live NYC Open Data catalog used by the package.

```{r nyc-list-datasets}
catalog <- nyc_list_datasets()

catalog |>
  filter(grepl("collision", name, ignore.case = TRUE)) |>
  select(key, uid, name)
```

The output includes columns such as the dataset title, description, and link to the source. The most important pieces are the key **and** uid. You need **either** in order to use the `nyc_pull_dataset()` function. You can put **either** the key value or uid value into the `dataset =` filter inside of `nyc_pull_dataset()`.

For instance, if we want to pull the the dataset `Motor Vehicle Collisions - Crashes`, we can use either of the methods below:

```{r nyc-311-pull}
nyc_motor_vehicle_collisions_data <- nyc_pull_dataset(
  dataset = "h9gi-nx95", limit = 2)

nyc_motor_vehicle_collisions_data <- nyc_pull_dataset(
  dataset = "motor_vehicle_collisions_crashes", limit = 2)
```

No matter if we put the `uid` or the `key` as the value for `dataset =`, we successfully get the data!

### The `nyc_any_dataset()` function

The easiest workflow is to use `nyc_list_datasets()` together with `nyc_pull_dataset()`. Because the catalog is retrieved dynamically from NYC Open Data, many newly published datasets are automatically available through the package.

In the event that you have a particular dataset you want to use in R that is not in the list, you can use the `nyc_any_dataset()`. The only requirement is the dataset’s API endpoint (a URL provided by the NYC Open Data Portal).

NYC Open Data JSON endpoints typically follow this structure:

`https://data.cityofnewyork.us/resource/<dataset_uid>.json`

For example, the Motor Vehicle Collisions dataset has the UID `"h9gi-nx95"`,
so its JSON endpoint becomes:

`https://data.cityofnewyork.us/resource/h9gi-nx95.json`

Here are the steps to get it:

1. On the NYC Open Data Portal, go to the dataset you want to work with.
2. Click on "Export" (next to the actions button on the right hand side).
3. Click on "API Endpoint".
4. Click on "SODA2" for "Version".
4. Copy the API Endpoint.

Below is an example of how to use the `nyc_any_dataset()` once the API endpoint has been discovered, that will pull the same data as the `nyc_pull_dataset()` example:

```{text}
nyc_motor_vehicle_collisions_data <- nyc_any_dataset(json_link = "https://data.cityofnewyork.us/resource/h9gi-nx95.json", limit = 2)
```

### Rule of Thumb

While both functions provide access to NYC Open Data, they serve slightly different purposes.

In general:

- Use `nyc_pull_dataset()` when the dataset is available in `nyc_list_datasets()`
- Use `nyc_any_dataset()` when working with datasets outside the catalog

Together, these functions allow users to either quickly access the datasets or flexibly query any dataset available on the NYC Open Data portal.

## Real World Example

NYC has a population of almost 8.5 million people, and while there are a lot of people taking public transportation, there are still many drivers. Unfortunately, there are sometimes crashes that take place, and all collision data are contained in the dataset, [found here](https://data.cityofnewyork.us/Public-Safety/Motor-Vehicle-Collisions-Crashes/h9gi-nx95/about_data). In R, the `nycOpenData` package can be used to pull this data directly.

By using the `nyc_pull_dataset()` function, we can gather recent motor vehicle collision records from New York City and filter based on any columns available in the dataset.

Let's take an example of the last 2 requests from the borough Brooklyn. The `nyc_pull_dataset()` function can filter based off any of the columns in the dataset. To filter, we add `filters = list()` and put whatever filters we would like inside. From our `colnames` call before, we know that there is a column called "borough" which we can use to accomplish this.

```{r filter-brooklyn}

brooklyn_collisions <- nyc_pull_dataset(dataset = "h9gi-nx95",limit = 2, timeout_sec = 90, filters = list(borough = "BROOKLYN"))
brooklyn_collisions

# Checking to see the filtering worked
brooklyn_collisions |>
  distinct(borough)
```

Success! From calling the `brooklyn_collisions` dataset we see there are only 2 rows of data, and from the `distinct()` call we see the only borough featured in our dataset is BROOKLYN.

One of the strongest qualities this function has is its ability to filter based off of multiple columns. Let's put everything together and get a dataset of the last *50* collisions in Brooklyn involving a Sedan.

```{r filter-brooklyn-nypd}
# Creating the dataset
brooklyn_sedan <- nyc_pull_dataset("h9gi-nx95", limit = 50, timeout_sec = 90, filters = list(vehicle_type_code1 = "Sedan", borough = "BROOKLYN"))

# Calling head of our new dataset
brooklyn_sedan |>
  slice_head(n = 6)

# Quick check to make sure our filtering worked
brooklyn_sedan |>
  summarize(rows = n())

brooklyn_sedan |>
  distinct(vehicle_type_code1)

brooklyn_sedan |>
  distinct(borough)
```

We successfully created a dataset containing the 50 most recent Brooklyn collisions involving a Sedan.

Advanced users may also provide raw SoQL queries through the `where` argument in `nyc_pull_dataset()`.

### Mini analysis

Now that we have successfully pulled the data and have it in R, let's do a mini analysis on using the `contributing_factor_vehicle_1` column, to figure out what are the main reasons for the collisions.

To do this, we will create a bar graph of the contributing factors associated with the collisions.

```{r compaint-type-graph, fig.alt="Bar chart showing the frequency of collision contributing factors in Brooklyn involving a Sedan.", fig.cap="Bar chart showing the frequency of collision contributing factors in Brooklyn involving a Sedan.", fig.height=5, fig.width=7}
# Visualizing the distribution, ordered by frequency
brooklyn_sedan |>
  count(contributing_factor_vehicle_1) |>
  ggplot(aes(
    x = n,
    y = reorder(contributing_factor_vehicle_1, n)
  )) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Top 50 Collisions in Brooklyn Involving a Sedan",
    x = "Number of Collisions",
    y = "Contributing Factor"
  )
```

This graph shows us not only which contributing factors appeared most frequently,
but also how often each factor occurred in the dataset.

## Summary

The `nycOpenData` package serves as a robust interface for the NYC Open Data portal, streamlining the path from raw city APIs to actionable insights. By simplifying common data acquisition tasks—such as filtering, downloading,
and optional type coercion—it allows users to focus on analysis rather than data engineering.

As demonstrated in this vignette, the package provides a seamless workflow for targeted data retrieval, automated filtering, and rapid visualization.

## How to Cite

If you use this package for research or educational purposes, please cite it as follows:

Martinez C (2026). nycOpenData: Convenient Access to NYC Open Data API Endpoints. R package version 0.2.3, <https://github.com/ropensci/nycOpenData>.
